Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2015-12-18 paper_mnras.tex
version for collabs

\title
[Testing Quasar Unification]{
% Modelling quasar outflows:
% clumpy winds and broad emission lines
Testing Quasar Unification: Radiative Transfer in Clumpy Winds
%Unifying quasars with clumpy wind models
}

%\author[Matthews et al.]
\author[Matthews et al.]{
\parbox[t]{\textwidth}{
James~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, Christian~Knigge$^1$,
Nick~Higginbottom$^1$, Knox~S.~Long$^2$, Stuart~A.~Sim$^3$ and Sam~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

\date{\today}
%\\
%Supervisor: Prof. Christian Knigge\\
%{\sl School of Physics \& Astronomy, University of Southampton,
%  Southampton, SO17 1BJ, UK}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          ABSTRACT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\maketitle
\begin{abstract} 
% Broad absorption lines (BALs) in the ultraviolet 
% are seen in $\sim20\%$ of quasi-stellar objects (QSOs). 
% Blue-shifted broad absorption lines (BALs) are the most direct evidence of 
% accretion disc `winds' in such systems; mass loaded outflows
% emanating from the disc that may be driven by line forces or
% magnetic processes. 
% Various unification schemes for
% quasars and luminous active galactic nuclei (AGN) have proposed
% that the broad emission line region is roughly cospatial
% with broad absorption line (BAL) gas and much of the phenomenology of luminous AGN
% can be explained by a simple geometrical picture involving an accretion
% disc and associated outflow. Here, we test this paradigm by 
% utilising our state-of-the-art radiative transfer code to produce synthetic spectra
% from simple biconical disc wind models. 
% In particular, we expand on our previous 
% work in which a benchmark model for BAL quasars was produced. 
% We have conducted a limited parameter search with the aim of unifying quasar phenomenology,
% and arrive at an improved model. The grid is now publicly available.
% We find that a simple treatment of clumping (`microclumping') 
% allows for a more realistic X-ray luminosity in the model by lowering the 
% ionization parameter. We examine the X-ray properties of this new model
% and find good agreement with existing X-ray samples of AGN and QSOs.
% We find that the dense, X-ray heated wind 
% produces strong H recombination and collisionally excited resonance 
% line emission to emerge at the low inclination angles, 
% which represent quasars within this unification scenario.
% % We also treat Hydrogen using a `macro-atom' approach in order to 
% % examine the effect of recombination on Hydrogen emission lines, and
% % this results in significant line emission in Lyman~$\alpha$ and the Balmer series.
% However, we are unable to reproduce the remarkably similar of line-to-continuum
% ratios between BAL and non-BAL quasar composites.
% This is due to a fundamental constraint arising
% from the anisotropy of emission from a classical thin disc. 
% We briefly explore the effect of relativistic beaming, gravitational redshift and 
% light bending on the angular distribution of disc continuum emission. 
% We find that these general relativistic effects do cause the disc to
% emit slightly more isotropically, but this is not yet sufficient to produce a
% self-consistent model. We discuss wind reprocessing as a potential solution.
% Overall, our work suggests that geometric unification
% involving an accretion disc wind is a promising scenario, but our results 
% pose a number of difficult challenges to such a model.
% Determining the true geometry of ADWs and uncovering the true disc spectral 
% (and angular) energy distribution are key next stpng if we are to build up a 
% holistic picture of the quasar population.
Various unification schemes have been proposed to interpret the complex phenomenology 
of quasars and luminous active galactic nuclei (AGN) in terms of  a simple axisymmetric 
picture involving a central black hole, an accretion disc and an associated outflow.
Here, we continue our tests of  this paradigm by comparing quasar spectra to 
synthetic spectra of simple biconical disc wind models, produced with our state-of-the-art 
Monte Carlo radiative transfer and photoionization code PYTHON.  
Previously, we have shown that we could produce synthetic spectra 
resembling those of observed broad absorption line (BAL) quasars 
for plausible wind models, but only if the X-ray luminosity was limited to about  
10$^{43}$~erg~s$^{-1}$. Here, we investigate the  degree to which
clumping of the outflow allows us to 
produce synthetic spectra in the rest-frame ultraviolet (UV) 
that have the characteristics of quasars with X-ray luminosities as 
large as 10$^{45}$ erg s$^{-1}$. 
We conclude, perhaps not surprisingly, that clumping does maintain the ionization 
state of the wind necessary for strong BAL features in the rest-frame UV at more 
realistic X-ray luminosities. We examine the X-ray properties of our fiducial clumped model
and find good agreement with existing X-ray samples of AGN and quasars.
In addition, the dense, X-ray heated wind produces strong recombination and collisionally 
excited line emission in, e.g., \la and \civ\, to emerge at the low inclination, 
`Type 1 quasar-like' angles. At the highest inclinations, the synthetic spectra 
possess prominent \mg\ and \al\ BALs, the absorption features seen in LoBAL quasars. 
Despite these successes, we are unable to reproduce all the emission lines seen in
quasar spectra with the correct equivalent width 
ratios in any of the models in our initial simulation grid,
and we find an angular dependence of emission line equivalent width despite the 
remarkable similarities in the observed spectral properties of BAL and non-BAL quasars.
Overall, our work suggests that BAL outflows can produce significant broad emission
lines and that accretion disc winds naturally reproduce the range of
ionization states seen in AGN, but a full quantitative agreement 
with quasar and BALQSO properties is not forthcoming.
% the remarkably uniform emission line properties 
% seen in BAL and non-BAL quasar composites, as we find a marked increase
% in emission line equivalent widths at high inclination `BALQSO-like' angles. 
% This is due to a fundamental constraint arising
% from the anisotropy of emission from a classical thin disc. 
% Overall, our work suggests that BAL outflows can produce significant broad emission
% lines and that accretion disc winds naturally reproduce the range of
% ionization states seen in AGN.
% This suggests that geometric unification
% involving an accretion disc wind is a promising scenario, but our results 
% pose a specific challenge to models in which an equatorial outflow
% rises from a limb-darkened, foreshortened accretion disc.
% {\bf There is confusion in the abstract, at least, in what is the model and what is the synthetic spectra, as well as whether there is one model or many models.  We need to try to be clear.  The model should be a clumped biconical outflow.  There are then particular instantiations of the model. And then there are synthetic spectra}
\end{abstract}

%\begin{keywords}
%AGN: outflows
%\end{keywords}

@@

Implementing a treatment of clumping is challenging, for
two main reasons. First, the physical scale lengths and density contrasts 
in AGN outflows are not well-constrained from observations.  
Second, there are significant computational difficulties associated with adequately
resolving and realistically modelling a series of small scale, high density
regions with a MCRT code. 
% Second, the addition of multiple additional degrees of
% freedom in the model results in significantly wider parameter space.
% Unfortunately, the physical scale lengths and density contrasts 
% associated with these parameters are not well-constrained from observations.  
% {\bf KSL: Reverse the order of these difficulties.  The most important is that it is not clear what kind of clumping an AGN wind has; the computational problem is a secondary one JM: Done..}
Given the lack of knowledge about the actual type of clumping, we have adopted
a simple approximation
used successfully in stellar wind modelling, known as 
{\em microclumping} \citep[e.g.][]{hamann1998,hilliermiller1999,hamann2008}. 
The underlying assumption of microclumping is that clump sizes are much smaller than the 
typical photon mean free path, and thus the clumps are 
both geometrically and optically thin. This approach 
allows one to introduce a `volume filling factor', $f_V$. 
The intra-clump medium is assumed to be a vacuum, so the 
density of the clumps is then multiplied by the ``density enhancement'' 
$D=1/f_V$. Opacities, $\kappa$, and emissivities, $\epsilon$, 
can then be expressed as 
\begin{equation}
\kappa = f_V \kappa_C(D);~~\epsilon = f_V \epsilon_C(D).
\end{equation}
Here the subscript $C$ denotes that the quantity is calculated using the 
enhanced density in the clump. The resultant effect is that, {\em for fixed temperature},
 processes that are linear in density, such as electron scattering, are unchanged, 
as $f_V$ and $D$ will cancel out. However, any quantity that scales with the square of density, 
such as collisional excitation or recombination, will increase by a factor of $D$.

@@

\begin{itemize}
\item UV absorption lines 
with $BI > 0$ at $\sim20\%$ of viewing angles (e.g. Knigge et al. 2010)
\item Line emission emerging at low inclinations, with $EW\sim40\AA$ in \civ\ \citep[e.g. ][]{shen2011}
\item H recombination lines with $EW\sim50\AA$ in \la\ \citep[e.g. ][]{shen2011}
\item  \mg\ and \al\ (LoBAL) absorption features with $BI > 0$ at a subset of 
BAL viewing angles
\item Verisimilitude with quasar composite spectra.
\end{itemize}
Here $BI$ is the `Balnicity Index' (Weymann et al. 1991), given by
\begin{equation}
BI = \int^{25000}_{3000} \left( 1 - \frac{f(v)}{0.9} \right) dv.
\end{equation}
The constant $C=0$ everywhere, unless the normalized flux
has satisfied $f(v)<0.9$ continuously for at least $2000$~km~s$^{−1}$, 
whereby $C$ is set to $1$.

% the following criteria:
% \begin{itemize}
% \item Do the synthetic spectra exhibit UV absorption lines 
% with $BI > 0$ at $\sim20\%$ of viewing angles (e.g. Knigge et al. 2010)?
% \item Do the synthetic spectra show  line emission emerge at low inclinations, with $EW\sim40\AA$ in \civ \citep[e.g. ][]{shen2011}?
% \item Do H recombination lines appear in the spectra, $EW\sim50\AA$ in \la 
% \citep[e.g. ][]{shen2011}?
% \item Do the spectra produce \mg\ and \al\ (LoBAL) absorption 
% features with $BI > 0$ at a small subset of BAL angles?
% \item Does the model spectra resemble quasar composite spectra?
% \end{itemize}
% Where $BI$ is the `Balnicity Index' (Weymann et al. 1991), given by
% \begin{equation}
% BI = \int^{25000}_{3000} \left( 1 - \frac{f(v)}{0.9} \right) dv.
% \end{equation}
% The constant $C=0$ everywhere, unless the normalized flux
% has satisfied $f(v)<0.9$ continuously for at least $2000$~km~s$^{−1}$, 
% whereby $C$ is set to $1$.

@@

% {\bf I still think we need to avoid using model to be a generic term to describe the outflow and the spectra.  If you mean the spectra use "model spectra" or "synthetic spectra'.  Note - the whole question of the model grid and the selection of one specific instantiation is not dealt with well.}

@@

The total line luminosity also increases dramatically compared to H13. 
This is primarily due to the higher X-ray luminosity increasing the bound-free
heating in the plasma, which avoids over-ionization due to clumping.
This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the incident X-rays.
The radii of these emitting regions is important,
and can be compared to observations. The line luminosities, $L$,
shown in the figure correspond to the luminosity in $\LUM$ of photons
escaping the Sobolev region for each line. 
% This is equivalent to $\beta_{ul} n_u A_{ul}$, where the three quantities
% represent the Sobolev escape probability, upper level number density and 
% Einstein A coefficient for the line. 
As shown in figure~\ref{fig:wind},
the \civ\ line in the fiducial model is typically formed between 
$100-1000~r_G$ ($\sim10^{17}-10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civ\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/uvspec_new.png}
\caption
{
Synthetic spectra at four viewing angles for the fiducial model. At 
$20^\circ$ and $60^\circ$ we show a comparison to an SDSS quasar composite
from Recihard et al. (2003). At $73^\circ$ and $76^\circ$ we show a comparison to
an {\sl HST} STIS spectrum of the high BALnicity BALQSO 
PG0946+301 (Arav et al. 2000), and an SDSS spectrum of the LoBAL quasar 
SDSS J162901.63+453406.0, respectively. The dotted line shows a disc
only continuum to show the effect of the outflow on the continuum level. 
All the spectra are scaled to the model flux at $2000$\AA, expect for the 
{\sl HST} STIS spectrum of PG0946+301, which is scaled using a continuum fit
due to the incomplete wavelength coverage.
% HST STIS spectrum of PG0946+301 (Arav et al. 2000)
}
\label{fig:uvspec}
\end{figure*}

@@

@@

The UV spectrum is characterised by strong BAL 
profiles at high inclinations ($> 70^\circ$). 
This highlights the first success of our model: 
clumping allows the correct ionization state 
to be maintained in the presence of strong X-rays, 
resulting in large resonance line opacities. 
At the highest inclinations, the 
cooler, low ionization material at the base of the wind
starts to intersect the line of sight. This produces 
multiple absorption lines in species such as \mg,
\al\ and Fe~\textsc{ii}. The potential links to LoBALQSOs and 
FeLoBALQSOs are discussed in section 2.4.
% Clearly, the absorption lines in the 
% synthetic spectra at BALQSO-like angles do not match the composites; this
% is {\em not} due to a failure of our model, but rather because a geometric mean
% BALQSO composite spectrum tends to wash out the broad absorption features due to the wide
% range of absorption characteristics. To demonstrate this, we show a comparison 
% to a {\sl Hubble Space Telescope} STIS spectrum of the high BALnicity BALQSO PG0946+301 
% (Arav et al. 2000) in figure \ref{fig:stis}.

The high ionization BAL profiles are often saturated, and the location in velocity space
of the strongest absorption in the profile varies with inclination.
At the lowest inclination BAL sight lines, the strongest absorption occurs at the red edge,
whereas at higher inclinations (and for the strongest BALs)
the trough has a sharp edge at the terminal velocity.
This offers one potential explanation for the wide range of BALQSO absorption
line shapes (see e.g. Trump et al. 2006; Knigge et al 2008, Filiz Ak et al. 2014).
In addition, the line profile shape is strongly dependent 
on the density, ionization and velocity 
profiles intersected by the line of sight. Thus, small tweaks of the velocity
law and angular distributions of streamlines can dramatically alter
the shape of the line.

Non-black saturation is observed in the absorption troughs of BALQSOs \citep{arav1999b,arav1999a}.
This is usually explained either as  partial covering of the continuum
source or by scattered contributions to the BAL troughs, necessarily
from an opacity source not co-spatial with the BAL forming region.
The scattered light explanation is supported by spectropolarimetry results
\citep{lamy2000}. Our spectra do not show nonblack saturation.
Instead, we find black, saturated troughs at angles $i > 73^\circ$, and the BALs
are non-saturated at lower inclinations. The reasons for this are inherent in the construction of our model. 
First, the microclumping assumption does not allow for 
porosity in the wind, meaning that it does not naturally produce
a partial covering absorber. To allow this, an alternative approach
such as {\em macroclumping} would be required \citep[e.g.][]{hamann2008,surlan2012}.
Second, our wind does not have a significant scattering 
contribution along sightlines which do not pass through the BAL region,
meaning that any scattered component to the BAL troughs is absorbed by line opacity.
This suggests that either the scattering cross-section of the wind must
be increased (with higher mass loss rates or covering factors), or 
that an additional source of electron opacity is required, potentially
in a polar direction above the disc.  
% {\bf KSL: it sounds like what you are suggesting is a low density inner flow, our domain 2 stuff actually permits ... or indeed one might be able to produce this within SV is you adopted a different mass loss with radius assumption. JM: I am being deliberately 
% vague about what I am suggesting. I am saying there are three physical
% possibilities: either wind itself causes this scattering contribution, which doesn't happen in our model, or you have partial covering, or you have some additional scatterer.}

@@

We find significant collisionally excited line emission emerges
at low inclinations in the synthetic spectra, particular in the \civ\ and \nv\
lines. The improved treatment of recombination also results in a strong \la\ line. 
In the context of unification, this is a promising result, 
and shows that a biconical wind can produce significant 
emission at `quasar-like' angles. To demonstrate this further,
we show line luminosities and monochromatic continuum luminosities
from the synthetic spectra in table~\ref{line_lums}. These are compared to
mean values from a subsample of the SDSS DR7 quasar catalog (Shen et al. 2011) 
with BH mass and Eddington fraction estimates similar to the fiducial model values 
(see caption). The spectra do not contain the strong 
C~\textsc{iii}]~1909\AA\, line seen in the quasar composite spectra,
because semi-forbidden (intercombination)
lines are not included in our modelling.
% This is because we do not fully solve the C level populations,
% and thus the level populations do not accurately reflect those
% as a full macro-atom with a 
% full collisional rates between forbidden
% or semi-forbidden transitions, as would be required.
The critical density of the C~\textsc{iii}]~1909\AA\, line 
is $n_e\sim10^{9.5}~\rm{cm^{-3}}$ \citep{wei1988}, which is higher than much of the 
outer portion of our wind. 
We therefore expect a model with these 
parameters to produce a C~\textsc{iii}]~1909\AA\ line
with a proper treament. We note, however, the the presence of
this line in the spectrum of quasars does {\em not} prohibit denser material having a significant contribution to the spectrum.

@@

The shapes and widths of these lines
match the composites fairly well, and the integrated 
line luminosities are significant when compared to observations.
However, there are a number of properties of the synthetic spectra
that are at odds with the observations. First, we find that the ratios of the 
equivalent width of the \la\ and \mg\ lines
to \civ\ is much lower than in the composite spectra. This problem 
has also been seen in simpler photoionization models for the BLR \citep{netzer1990}.
It may be that a larger region of very dense ($n_e\sim10^{10}$cm$^{-3}$) 
material is needed, this could correspond to a disc atmosphere or 
`transition region' (see e.g. Murray et al. 1995; Knigge et al. 1997). 
While modest changes to geometry may permit this, the initial grid search 
did not find a parameter space in which the \la\ or \mg\ equivalent widths
were significantly higher. Second, we find that equivalent widths increase
with inclinations, to the extent that significantly denser
models can match the line EWs fairly well at low inclinations but will then
possess overly strong red wings to the BAL P-Cygni profiles at high inclinations
(see also figure~\ref{fig:lobal}).
This angular dependence of EW is expected from a model in which the BLR radiates
more isotropically than the disc, which in this case is foreshortened and limb darkened.
It appears to contradict observations, which show remarkably uniform emission
line properties in quasars and BALQSOs \citep{weymann1991,dipompeo2012b}. 
The angular distribution of the disc 
continuum and line emission
is clearly crucially important in determining the emergent broad 
line equivalent widths, as suggested by, e.g., the analysis of \cite{risaliti2011}. 
We shall explore this question further in a future study.  
% However, the line-to-continuum ratios 
% at low inclinations in our model are significantly weaker than the quasar 
% composites. Increasing the density of the outflow, by altering the mass 
% loss rate  or velocity law, can produce more line emission.
% However, the red wing of the BAL profiles is generally stronger than 
% seen in BALQSO spectra and composites. This suggests a fairly 
% fundamental problem with a geometric unification model in which 
% The line equivalent widths at high inclinations are
% significantly affected by disc foreshortening and limb darkening.
% The angular distribution of the disc radiation is clearly
% crucially important in determining the emergent line ratios.  
% {\bf KSL: This is a bit oddly explained. Our problem is at low inclination.  Note that I am a bit concerned with Fig 3; the model spectra and the observations look very different.}

% {\bf I have to say I find this hard to understand.  Are you trying to say that in CVs which we are know are thin disks with winds, the EW of the red-wing of emission lines increases with inclination angle ... which I think may be true, and so one would expect that in a unification model for AGN  it should do the same?  And the fact that it does not, suggests that something else is going on in AGN.  I also still think that you should move this discussion later in the paper and not bury it here.}

@@

% JM XXX -- only include this if it's true!
% Our models slightly over-predict the emergent X-ray luminosity at BAL angles, 
% although we are limited by poor sample sizes. 
% % It is possible that this is due to our wind being overly optically thick to 
% % electron scattering at angles which look through the wind. 
% If BALQSOs were {\em intrinsically} 
% X-ray weak \citep[as suggested by, e.g.][]{morabito2013},
% our isotropic assumption for the X-ray source would be incorrect. 
% A polar-biased X-ray source would result in a lower clumping factor being
% required in our model. Our specific wind
% prescription will also affect the opacities, densities and resultant
% ionization structure, which can change the absorption 
% characteristics and resultant luminosities.
% Nevertheless, our input X-ray spectrum
% now reproduces the X-ray properties of a luminous quasar as an output,
% and at least some BAL angles match the observations.
% This satisifies the first-order requirement for the X-ray properties of 
% a unified quasar model.

>>>>

\title
[Testing Quasar Unification: Radiative Transfer in Clumpy Winds]{
% Modelling quasar outflows:
% clumpy winds and broad emission lines
Testing Quasar Unification: Radiative Transfer in Clumpy Winds
%Unifying quasars with clumpy wind models
}

\author[Matthews et al.]{
\parbox[t]{\textwidth}{
James~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, Christian~Knigge$^1$,
Nick~Higginbottom$^1$, Knox~S.~Long$^2$, Stuart~A.~Sim$^3$ and Sam~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          ABSTRACT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\maketitle
\begin{abstract} 
Various unification schemes have been proposed to interpret the complex phenomenology 
of quasars and luminous active galactic nuclei (AGN) in terms of  a simple axisymmetric 
picture involving a central black hole, an accretion disc and an associated outflow.
Here, we continue our tests of  this paradigm by comparing quasar spectra to 
synthetic spectra of simple biconical disc wind models, produced with our state-of-the-art 
Monte Carlo radiative transfer and photoionization code PYTHON.  
Previously, we have shown that we could produce synthetic spectra 
resembling those of observed broad absorption line (BAL) quasars 
for plausible wind models, but only if the X-ray luminosity was limited to about  
10$^{43}$~erg~s$^{-1}$. Here, we investigate the  degree to which
clumping of the outflow allows us to 
produce synthetic spectra in the rest-frame ultraviolet (UV) 
that have the characteristics of quasars with X-ray luminosities as 
large as 10$^{45}$ erg s$^{-1}$. 
We conclude, perhaps not surprisingly, that clumping does maintain the ionization 
state of the wind necessary for strong BAL features in the rest-frame UV at more 
realistic X-ray luminosities. We examine the X-ray properties of our fiducial clumped model
and find good agreement with existing X-ray samples of AGN and quasars.
In addition, the dense, X-ray heated wind produces strong recombination and collisionally 
excited line emission in, e.g., \la\ and \civline, to emerge at the low inclination, 
`Type 1 quasar-like' angles. At the highest inclinations, the synthetic spectra 
possess prominent \mg\ and \al\ BALs, the absorption features seen in LoBAL quasars. 
Despite these successes, we are unable to reproduce all the emission lines seen in
quasar spectra with the correct equivalent width 
ratios in any of the models in our initial simulation grid,
and we find an angular dependence of emission line equivalent width despite the 
remarkable similarities in the observed spectral properties of BAL and non-BAL quasars.
Overall, our work suggests that a biconical wind model
reproduces much of the qualitative behaviour expected from a unified model. 
However, we cannot yet provide quantitative matches with quasar properties at all 
viewing angles; whether disc winds can successfully unify quasars is therefore 
still an open question.
\end{abstract}

% \begin{keywords}
% galaxies: active; accretion, accretion discs; (galaxies:) quasars: emission lines;
% (galaxies:) quasars: absorption lines
% \end{keywords}

@@

Implementing a treatment of clumping is challenging, for
two main reasons. First, the physical scale lengths and density contrasts 
in AGN outflows are not well-constrained from observations.  
Second, there are significant computational difficulties associated with adequately
resolving and realistically modelling a series of small scale, high density
regions with a MCRT code. 
Given the lack of knowledge about the actual type of clumping, we have adopted
a simple approximation
used successfully in stellar wind modelling, known as 
{\em microclumping} \citep[e.g.][]{hamann1998,hilliermiller1999,hamann2008}. 
The underlying assumption of microclumping is that clump sizes are much smaller than the 
typical photon mean free path, and thus the clumps are 
both geometrically and optically thin. This approach 
allows one to introduce a `volume filling factor', $f_V$. 
The intra-clump medium is assumed to be a vacuum, so the 
density of the clumps is then multiplied by the ``density enhancement'' 
$D=1/f_V$. Opacities, $\kappa$, and emissivities, $\epsilon$, 
can then be expressed as 
\begin{equation}
\kappa = f_V \kappa_C(D);~~\epsilon = f_V \epsilon_C(D).
\end{equation}
Here the subscript $C$ denotes that the quantity is calculated using the 
enhanced density in the clump. The resultant effect is that, {\em for fixed temperature},
 processes that are linear in density, such as electron scattering, are unchanged, 
as $f_V$ and $D$ will cancel out. However, any quantity that scales with the square of density, 
such as collisional excitation or recombination, will increase by a factor of $D$.

@@

\begin{itemize}
\item UV absorption lines 
with $BI > 0$ at $\sim20\%$ of viewing angles (e.g. Knigge et al. 2010)
\item Line emission emerging at low inclinations, with $EW\sim40$\AA in \civline\ \citep[e.g. ][]{shen2011}
\item H recombination lines with $EW\sim50$\AA in \la\ \citep[e.g. ][]{shen2011}
\item  \mg\ and \al\ (LoBAL) absorption features with $BI > 0$ at a subset of 
BAL viewing angles
\item Verisimilitude with quasar composite spectra.
\end{itemize}
Here $BI$ is the `Balnicity Index' (Weymann et al. 1991), given by
\begin{equation}
BI = \int^{25000}_{3000} \left( 1 - \frac{f(v)}{0.9} \right) dv.
\end{equation}
The constant $C=0$ everywhere, unless the normalized flux
has satisfied $f(v)<0.9$ continuously for at least $2000$~km~s$^{−1}$, 
whereby $C$ is set to $1$.

@@

@@

The total line luminosity also increases dramatically compared to H13. 
This is primarily due to the higher X-ray luminosity increasing the bound-free
heating in the plasma, which avoids over-ionization due to clumping.
This line emission typically emerges on the edge of the wind
nearest the central source. The location of the line emitting regions
is dependent on the ionization state, as well as the incident X-rays.
The radii of these emitting regions is important,
and can be compared to observations. The line luminosities, $L$,
shown in the figure correspond to the luminosity in $\LUM$ of photons
escaping the Sobolev region for each line. 
% This is equivalent to $\beta_{ul} n_u A_{ul}$, where the three quantities
% represent the Sobolev escape probability, upper level number density and 
% Einstein A coefficient for the line. 
As shown in figure~\ref{fig:wind},
the \civline\ line in the fiducial model is typically formed between 
$100-1000~r_G$ ($\sim10^{17}-10^{18}~\rm{cm}$).
This is in rough agreement with the reverberation mapping 
results of Kaspi (2000) for the $2.6\times10^{9} M_\odot$ quasar S5 0836+71,
and also compares favourably with microlensing measurements of the size of the
\civline\ emission line region in the BALQSO H1413+117 \citep{odowd2015}.

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/uvspec_new.png}
\caption
{
Synthetic spectra at four viewing angles for the fiducial model. At 
$20^\circ$ and $60^\circ$ we show a comparison to an SDSS quasar composite
from Recihard et al. (2003). At $73^\circ$ and $76^\circ$ we show a comparison to
an {\sl HST} STIS spectrum of the high BALnicity BALQSO 
PG0946+301 (Arav et al. 2000), and an SDSS spectrum of the LoBAL quasar 
SDSS J162901.63+453406.0, respectively. The dotted line shows a disc
only continuum to show the effect of the outflow on the continuum level. 
All the spectra are scaled to the model flux at $2000$\AA, expect for the 
{\sl HST} STIS spectrum of PG0946+301, which is scaled to $1350$\AA
due to the incomplete wavelength coverage.
% HST STIS spectrum of PG0946+301 (Arav et al. 2000)
}
\label{fig:uvspec}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=1.0\textwidth]{figures/sed.png}
\caption
{
Synthetic spectra at two viewing angles, 
this time in frequency space and including the optical band,
compared to the non-BAL SDSS quasar composite. The spectra are normalised to the flux at 
$2000$\AA, then an offset of 2 is applied per spectrum for clarity. 
}
\label{fig:sed}
\end{figure*}

@@

The UV spectrum is characterised by strong BAL 
profiles at high inclinations ($> 70^\circ$). 
This highlights the first success of our model: 
clumping allows the correct ionization state 
to be maintained in the presence of strong X-rays, 
resulting in large resonance line opacities. 
At the highest inclinations, the 
cooler, low ionization material at the base of the wind
starts to intersect the line of sight. This produces 
multiple absorption lines in species such as \mg,
\al\ and Fe~\textsc{ii}. The potential links to LoBALQSOs and 
FeLoBALQSOs are discussed in section 2.4.

The high ionization BAL profiles are often saturated, and the location in velocity space
of the strongest absorption in the profile varies with inclination.
At the lowest inclination BAL sight lines, the strongest absorption occurs at the red edge,
whereas at higher inclinations (and for the strongest BALs)
the trough has a sharp edge at the terminal velocity.
This offers one potential explanation for the wide range of BALQSO absorption
line shapes (see e.g. Trump et al. 2006; Knigge et al 2008, Filiz Ak et al. 2014).
% In addition, the line profile shape is strongly dependent 
% on the density, ionization and velocity 
% profiles intersected by the line of sight. Thus, small tweaks of the velocity
% law and angular distributions of streamlines can dramatically alter
% the shape of the line.

Non-black saturation is observed in the absorption troughs of BALQSOs \citep{arav1999b,arav1999a}.
This is usually explained either as  partial covering of the continuum
source or by scattered contributions to the BAL troughs, necessarily
from an opacity source not co-spatial with the BAL forming region.
The scattered light explanation is supported by spectropolarimetry results
\citep{lamy2000}. Our spectra do not show nonblack saturation.
Instead, we find black, saturated troughs at angles $i > 73^\circ$, and the BALs
are non-saturated at lower inclinations. The reasons for this are inherent in the construction of our model. 
First, the microclumping assumption does not allow for 
porosity in the wind, meaning that it does not naturally produce
a partial covering absorber. To allow this, an alternative approach
such as {\em macroclumping} would be required \citep[e.g.][]{hamann2008,surlan2012}.
Second, our wind does not have a significant scattering contribution 
along sightlines which do not pass through the BAL region,
meaning that any scattered component to the BAL troughs is absorbed by line opacity.
This suggests that either the scattering cross-section of the wind must
be increased (with higher mass loss rates or covering factors), or 
that an additional source of electron opacity is required, potentially
in a polar direction above the disc. We note the scattering contribution
from plasma in polar regions is significant in some `outflow-from-inflow'
simulations \citep{KP09, simproga2012}.

@@

We find significant collisionally excited line emission emerges
at low inclinations in the synthetic spectra, particular in the \civ\ and \nv\
lines. The improved treatment of recombination also results in a strong \la\ line. 
In the context of unification, this is a promising result, 
and shows that a biconical wind can produce significant 
emission at `quasar-like' angles. To demonstrate this further,
we show line luminosities and monochromatic continuum luminosities
from the synthetic spectra in table~\ref{line_lums}. These are compared to
mean values from a subsample of the SDSS DR7 quasar catalog (Shen et al. 2011) 
with BH mass and Eddington fraction estimates similar to the fiducial model values 
(see caption). The spectra do not contain the strong 
C~\textsc{iii}]~1909\AA\, line seen in the quasar composite spectra,
because semi-forbidden (intercombination)
lines are not included in our modelling.

In figure~\ref{fig:sed}, we show an $F_{\nu}$ spectrum with broader waveband coverage
that includes the optical, showing that our synthetic spectra 
also exhibit \ha\ and \hb\ emission. 
In this panel, we include a low inclination and 
also a very high inclination 
spectrum, which looks underneath the wind cone. This model shows 
strong line emission with very similar widths and line ratios to the quasar composites,
although velocity projection effects mean the Balmer lines are double peaked, 
as seen  in `disc emitter' systems.
The increased line equivalent widths (EWs) are due to a weakened continuum at high inclination from wind attenuation and 
disc foreshortening / limb darkening. This effect also 
leads to a redder continuum slope, as seen in quasars, which is
due to Balmer continuum and Balmer and Fe~\textsc{ii} line emission.
The high inclination viewing angle is a physically motivated quasar model,
but does show that a model such as this can naturally reproduce quasar emission lines
if the emissivity of the wind is increased {\em with respect to the disc continuum}.
In addition, it neatly demonstrates how a stratified outflow can naturally
reproduce the range of ionization states seen in quasars.
% This is because we do not fully solve the C level populations,
% and thus the level populations do not accurately reflect those
% as a full macro-atom with a 
% full collisional rates between forbidden
% or semi-forbidden transitions, as would be required.
% The critical density of the C~\textsc{iii}]~1909\AA\, line 
% is $n_e\sim10^{9.5}~\rm{cm^{-3}}$ \citep{wei1988}, which is higher than much of the 
% outer portion of our wind. 
% We therefore expect a model with these 
% parameters to produce a C~\textsc{iii}]~1909\AA\ line
% with a proper treament. We note, however, the the presence of
% this line in the spectrum of quasars does {\em not} prohibit denser material having a significant contribution to the spectrum.

@@

Despite a number of successes, 
there are a some properties of the synthetic spectra
that are at odds with the observations. First, we find that the ratios of the 
EW of the \la\ and \mgline\ lines
to the EW of \civline\ is much lower than in the composite spectra. 
This problem has also been seen in simpler photoionization models for the 
BLR \citep{netzer1990}.
It may be that a larger region of very dense ($n_e\sim10^{10}$cm$^{-3}$) 
material is needed, which could correspond to a disc atmosphere or 
`transition region' (see e.g. Murray et al. 1995; Knigge et al. 1997). 
While modest changes to geometry may permit this, the initial grid search 
did not find a parameter space in which the \la\ or \mg\ EWs
were significantly higher (see section~\ref{sec:param_sens}. 
Second, we find that EWs increase
with inclination, to the extent that significantly denser
models can match the line EWs fairly well at low inclinations but will then
possess overly strong red wings to the BAL P-Cygni profiles at high inclinations
(see figures~\ref{fig:uvspec} and \ref{fig:sed}; also figure~\ref{fig:lobal}).
This angular dependence is expected from a model in which the BLR radiates
more isotropically than the disc, which in this case is foreshortened and limb darkened.
It appears to contradict observations, which show remarkably uniform emission
line properties in quasars and BALQSOs \citep{weymann1991,dipompeo2012b}. 
The angular distribution of the disc 
continuum and line emission
is clearly crucially important in determining the emergent broad 
line EWs, as suggested by, e.g., the analysis of \cite{risaliti2011}. 
We shall explore this question further in a future study.

@@
